---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PointedSDMs

<!-- badges: start -->
[![R-CMD-check](https://github.com/PhilipMostert/PointedSDMs/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PhilipMostert/PointedSDMs/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of PointedSDMs is to simplify the construction of integrated species distribution models (ISDMs) for large collections of heterogeneous data. It does so by building wrapper functions around [inlabru](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.13168), which uses the [INLA methodology](https://rss.onlinelibrary.wiley.com/doi/abs/10.1111/j.1467-9868.2008.00700.x) to estimate a class of latent Gaussian models.


## Installation

You can install the development version of PointedSDMs from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("PhilipMostert/PointedSDMs")
```

## Example

This is a basic example which shows you how to specify and run an integrated model, using three disparate datasets containing locations of the solitary tinamou.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r example}
library(PointedSDMs)
library(raster)

```

```{r data}

#Load data in

data("SolitaryTinamou")

projection <- CRS("+proj=longlat +ellps=WGS84")

species <- SolitaryTinamou$datasets

Forest <- SolitaryTinamou$covariates$Forest

crs(Forest) <- projection

mesh <- SolitaryTinamou$mesh
mesh$crs <- projection

```

```{r intModel, message = FALSE, warning = FALSE}

#Specify model -- here we run a model with one spatial covariate and a shared spatial field

model <- intModel(species, spatialCovariates = Forest, Coordinates = c('X', 'Y'),
                 Projection = projection, Mesh = mesh, responsePA = 'Present')

```

```{r runModel}

#Run the integrated model

modelRun <- runModel(model, options = list(control.inla = list(int.strategy = 'eb')))
summary(modelRun)

```

