---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PointedSDMs

<!-- badges: start -->

[![R-CMD-check](https://github.com/PhilipMostert/PointedSDMs/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PhilipMostert/PointedSDMs/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/PhilipMostert/PointedSDMs/branch/ChangingToR6/graph/badge.svg)](https://app.codecov.io/gh/PhilipMostert/PointedSDMs?branch=ChangingToR6)

<!-- badges: end -->

The goal of *PointedSDMs* is to simplify the construction of integrated species distribution models (ISDMs) for large collections of heterogeneous data. It does so by building wrapper functions around [inlabru](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.13168), which uses the [INLA methodology](https://rss.onlinelibrary.wiley.com/doi/abs/10.1111/j.1467-9868.2008.00700.x) to estimate a class of latent Gaussian models.

## Installation

You can install the development version of PointedSDMs from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("PhilipMostert/PointedSDMs")
```

## Package functionality

*PointedSDMs* includes a selection of functions used to streamline the construction of ISDMs as well and perform model cross-validation. The core functions of the package are:

| Function name  | Function description                                                                                          |
|-----------------|-------------------------------------------------------|
| `intModel()`   | Initialize and specify the components used in the integrated model.                                           |
| `blockedCV()`  | Perform spatial blocked cross-validation.                                                                     |
| `runModel()`   | Estimate the components of the integrated model.                                                              |
| `datasetOut()` | Perform dataset-out cross-validation, which calculates the impact individual datasets have on the full model. |

The function `intModel()` produces an [R6](https://github.com/r-lib/R6) object, and as a result there are various *slot functions* available to further specify the components of the model. These *slot functions* include:

| `intModel()` slot function   | Function description                                                                                                                                            |
|------------------|------------------------------------------------------|
| `` `.$plot()` ``             | Used to create a plot of the available data. The output of this function is an object of class [`gg`](https://github.com/tidyverse/ggplot2).                    |
| `` `.$addBias()` ``          | Add an additional spatial field to a dataset to account for sampling bias in unstructured datasets.                                                             |
| `` `.$updateFormula()` ``    | Used to update a formula for a process. The idea is to start specify the full model with `intModel()`, and then thin components per dataset with this function. |
| `` `.$updateComponents()` `` | Change or add new components used by [inlabru](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.13168) in the integrated model.            |
| `` `.$priorsFixed()` ``      | Change the specification of the prior distribution for the fixed effects in the model.                                                                          |
| `` `.$specifySpatial()` ``   | Specify the spatial field in the model using penalizing complexity (PC) priors.                                                                                 |
| `` `.$spatialBlock()` ``     | Used to specify how the points are spatially blocked. Spatial cross-validation is subsequently performed using `blockedCV()`.                                   |

## Example

This is a basic example which shows you how to specify and run an integrated model, using three disparate datasets containing locations of the solitary tinamou (*Tinamus solitarius)*.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r example}

library(PointedSDMs)
library(ggplot2)
library(raster)

```

```{r data}

#Load data in

data("SolitaryTinamou")

projection <- CRS("+proj=longlat +ellps=WGS84")

species <- SolitaryTinamou$datasets

Forest <- SolitaryTinamou$covariates$Forest

crs(Forest) <- projection

mesh <- SolitaryTinamou$mesh
mesh$crs <- projection

```

Setting up the model is done easily with `intModel()`, where we specify the required components of the model:

```{r intModel, message = FALSE, warning = FALSE}

#Specify model -- here we run a model with one spatial covariate and a shared spatial field

model <- intModel(species, spatialCovariates = Forest, Coordinates = c('X', 'Y'),
                 Projection = projection, Mesh = mesh, responsePA = 'Present')

```

We can also make a quick plot of where the species are located using `` `.$plot()` ``:

```{r plot, warning = FALSE, message = FALSE, fig.width=8, fig.height=5}

region <- SolitaryTinamou$region

model$plot(Boundary = FALSE) + gg(region)

```

We can estimate the parameters in the model using the `runModel()` function:

```{r runModel, warning = FALSE, message = FALSE}

#Run the integrated model

modelRun <- runModel(model, options = list(control.inla = list(int.strategy = 'eb')))
summary(modelRun)

```

*PointedSDMs* also includes generic predict and plot functions:

```{r predict_and_plot, warning = FALSE, message = FALSE, fig.width=8, fig.height=5}

predictions <- predict(modelRun, mesh = mesh,
                       mask = region, 
                       spatial = TRUE,
                       fun = 'linear')

plot(predictions)

```
