---
title: "Example of PointedSDMs for the solitary tinamou"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example of PointedSDMs for the solitary tinamou}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

```

## Introduction

This vignette illustrates the various functions of *PointedSDMs* by using three datasets of the solitary tinamou (*Tinamus solitarius) --* a species of ground bird found on the eastern side of Brazil. Due to package dependencies, this vignette is not run. However the data and *R* script are available such that the user may carry out inference.

```{r setup, warning = FALSE, message = FALSE}

library(PointedSDMs)
library(terra)
library(ggpolypath)
library(INLA)
library(ggplot2)

```

```{r safe, include = FALSE}

bru_options_set(inla.mode = "experimental")

```

### Load data

Firstly, we load in the datasets and objects required for this vignette. The *SolitaryTinamou* dataset attached to this package contains a `list` of four objects; for ease of use, we make new objects for the items in this `list`.

```{r load data}

data('SolitaryTinamou')
projection <- "+proj=longlat +ellps=WGS84"

covariates <- terra::rast(system.file('extdata/SolitaryTinamouCovariates.tif', 
                                      package = "PointedSDMs"))

datasets <- SolitaryTinamou$datasets
region <- SolitaryTinamou$region
mesh <- SolitaryTinamou$mesh


```

The first item is a `list` of three datasets: *eBird*, *Gbif* and *Parks*. The first two are `data.frame` objects containing only two variables: `X` and `Y` describing the latitude and longitude coordinates of the species location respectively. As a result of this, these two datasets are considered to be *present only* datasets in our integrated model.

The other dataset (*Parks*) is also a `data.frame` object. It contains the two coordinate variables present in the first two datasets, but contains two additional variables: `Present`, a binary variable describing the presence (*1*) or absence (*0*) of the species at the given coordinates, and `area` describing the area of the park. Since we have information on the presences and absences of the species in this dataset, we consider it a *presence absence* dataset.

`Region` is a `sf` object which give the boundary of the park containing the species; it was used in the *mesh* construction and for the plots in this vignette.

```{r look at data}

str(datasets)
class(region)

```

The next object is `covariates`, a `spatRaster` objects of the covariates (*Forest, NPP* and *Altitude*) describing the area of the parks. We stack these three objects together using the `stack` function, and then scale them.

```{r covariates, fig.width=8, fig.height=5}

covariates <- scale(covariates)
crs(covariates) <- projection
plot(covariates)

```

Finally we require a *Delaunay triangulated mesh* for the construction of the spatial field. A plot of the mesh used for this vignette is provided below.

```{r mesh, fig.width=8, fig.height=5}

plot(mesh)

```

### Base model

To set up an integrated species distribution model with `PointedSDMs`, we initialize it with the `intModel` function -- which results in an *R6* objects with additional slot functions to further customize the model. The base model we run for these data comprises of the spatial covariates and an intercept term for each dataset.

```{r set up base model, warning = FALSE, message = FALSE}

base <- intModel(datasets, spatialCovariates = covariates, Coordinates = c('X', 'Y'),
                 Projection = projection, responsePA = 'Present', Offset = 'area',
                 Mesh = mesh, pointsSpatial = NULL)

```

Using the `.$plot` function produces a *gg* object of the points used in this analysis by dataset; from this plot, we see that most of the species locations are found towards the eastern and south-central part of the park.

```{r data, fig.width=8, fig.height=5}

base$plot(Boundary = FALSE) + 
  geom_sf(data = st_boundary(region)) +
  ggtitle('Plot of the species locations by dataset')

```

In this model, we also include prior information for the *Forest* effect using `$priorsFixed`.

```{r priorsFixed}

base$priorsFixed(Effect = 'Forest', mean.linear = 0.5, prec.linear = 0.01)

```

To run the integrated model, we use the `fitISDM` function with the `data` argument as the object created with the `intModel` function above.

```{r run base model, warning = FALSE, message = FALSE}

baseModel <- fitISDM(data = base)
baseModel

```

### Model with spatial fields

#### Shared spatial field

Spatial fields are fundamental in our spatial species distribution models, and so we include them in the model by setting `pointsSpatial = TRUE` in `intModel`. Furthermore, we will remove the intercept terms by specifying `pointsIntercept = FALSE`

```{r set up model with fields, warning = FALSE, message = FALSE}


fields <- intModel(datasets, Coordinates = c('X', 'Y'),
                   Projection = projection, Mesh = mesh, responsePA = 'Present', 
                   Offset = 'area',
                   pointsIntercept = FALSE)

```

To specify the spatial field in the model, we use the slot function `$specifySpatial`. This in turn will call *R-INLA*'s `inla.spde2.pcmatern` function, which is used to specify penalizing complexity (PC) priors for the parameters of the field. If we had set `PC = FALSE` in this function, our shared spatial field would be specified with *R-INLA*'s `inla.spde2.matern` function.

```{r specifySpatial}

fields$specifySpatial(sharedSpatial = TRUE, prior.range = c(1,0.01), 
                      prior.sigma = c(0.75, 0.05))

```

We furthermore include an additional spatial field (deemed the *bias field*) for our citizen science *eBird* observations with the `$addBias` slot function.

```{r addBias}

fields$addBias('eBird')

```

Finally we run the integrated model, again with `fitISDM` but this time we specify options with *R-INLA*'s *empirical Bayes* integration strategy to help with computation time.

```{r run fields model, warning = FALSE, message = FALSE}

fieldsModel <- fitISDM(fields, options = list(control.inla = list(int.strategy = 'eb')))
fieldsModel

```

#### Copy field

If we would like to share the spatial fields across the linear predictors using *INLA*'s `copy` feature, we can specify `pointsSpatial = 'copy'` in `intModel()`:

```{r copy model}

copy <- intModel(datasets, Coordinates = c('X', 'Y'),
                 Projection = projection, Mesh = mesh, responsePA = 'Present', 
                 Offset = 'area', pointsSpatial = 'copy')

copy$changeComponents()

```

```{r run copy model}

copyModel <- fitISDM(copy, options = list(control.inla = list(int.strategy = 'eb')))
copyModel

```

### Prediction of the spatial fields

If we wanted to make predictions of the shared spatial random field across the map, we set `spatial = TRUE` in the generic `predict` function.

```{r predict spatial, warning = FALSE, message = FALSE}

spatial_predictions <- predict(fieldsModel, mesh = mesh,
                       mask = region, 
                       spatial = TRUE,
                       fun = 'linear')

```

And subsequently plot using the generic `plot` function.

```{r spatial, fig.width=8, fig.height=5}

plot(spatial_predictions)

```

However if we wanted to make predictions of the bias field, we would do this by setting `biasfield = TRUE`.

```{r predict bias, warning = FALSE, message = FALSE}

bias_predictions <- predict(fieldsModel, 
                    mesh = mesh, 
                    mask = region, 
                    biasfield = TRUE,
                    fun = 'linear')

```

```{r bias, fig.width=8, fig.height=5}

plot(bias_predictions)

```

### Dataset out cross-validation

The last function of interest is `datasetOut`, which removes a dataset out of the full model, and then calculates a cross-validation score with this reduced model. In this case, we try the function out by removing the *eBird* dataset.

```{r datasetOut, warning = FALSE, message = FALSE}

eBird_out <- datasetOut(model = fieldsModel, dataset = 'eBird')

```

```{r print datasetOut}

eBird_out

```
